Synthesis of Linear Functional Programs

### 2021-07-22 article/article.tex
Add focusing judgments, explanations, do todos

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesis must deal with two main
inherent sources of complexity -- search over the space of
valid programs, and interpreting user intent.

Synthesis is said to be \emph{type-driven} when it uses types as a form of program
specification and produces an expression whose type matches the specification.
Type-driven synthesis frameworks usually leverage rich types as a way to make
specifications more expressive and prune the valid programs search space, while
maintaining a ``familiar'' specification interface (types) for the user.
%
Richer type systems allow for more precise types, which can statically eliminate
various kinds of logical errors by making certain invalid program states
ill-typed (e.g.,~a ``null aware'' type system will ensure at compile-time that
you cannot dereference a null-pointer).
%
% However, they can also be a burden -- the whole point of these type systems is
% to ensure that \emph{fewer} programs are deemed well-typed, which can pose
% additional challenges to the development process.
%
For instance, the type $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow
\mathsf{Int}$ specifies a (curried) function that takes two integers and
produces an integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).  However,
the richer type $(x{:}\mathsf{Int}) \rightarrow (y{:}\mathsf{Int}) \rightarrow
\{z{:}\mathsf{Int} \mid z = x+y\}$ very precisely specifies a function that
takes integer arguments $x$ and $y$ and returns an integer $z$ that is the sum
of $x$ and $y$.

The focus of our work is on type-driven synthesis where specifications take the
form of linear types.
%
% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
% proof-theoretic foundations of linear types via the Curry-Howard
% correspondence with linear logic propositions, viewing program synthesis as
% proof search in linear logic.  (Said correspondence maps \emph{types} to
% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
%
Linear types constrain resource
usage in programs by \emph{statically} limiting the number of times certain
resources can be used during their lifetime (linear resources must be used
\emph{exactly once}).
%
They can be applied to resource-aware programming such as concurrent programming
(e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
(e.g.~Rust's ownership types), safely updating-in-place mutable
structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
% relacionar o nosso trabalho com futuras aplicações nestes campos

\paragraph{Contributions and Outline.} Despite their long-known potential
\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
and strong proof-theoretic
foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types combined with other advanced typing features
has generally been overlooked in the literature.  In this work we present a framework for synthesis
with linear types extended with recursive algebraic data types, parametric
polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?
We first introduce linear types as specifications
and outline the synthesis process, leveraging linearity, by example
(\S~\ref{sec:overview}). We then present the formal system driving the
synthesis (\S~\ref{sec:formal_system}) and describe the architecture
of our framework named \synname, examining technical details and key
implementation challenges (\S~\ref{sec:architecture}). Finally,
we evaluate our work through expressiveness and performance benchmarks
(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
%TODO: And appendice with...

@@

\paragraph{Core linear calculus.} Initially, we synthesize from
specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
\oplus, \bang$), such as linear logic theorems. Namely,
the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
the function \mintinline{haskell}{\a -> \b -> let c*d = b in a c d},
which takes an argument \mintinline{haskell}{a}, a
\emph{curried} linear function of type $A \lolli B \lolli C$,  and an
argument \mintinline{haskell}{b}, a \emph{linear
  pair} of an $A$ and a $B$ and produces a value of type $C$ by
deconstructing the pair and applying \mintinline{haskell}{a} to the
corresponding elements of the pair (i.e.,~\emph{uncurries} the
function \mintinline{haskell}{a}).

@@

%TODO: Polymorphism ? read todo below

\todo[inline]{Tens de explicar o que significa o $\bang$, see above.}
\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
used linearly (consumed), we can assure, e.g., the deconstruction of an
ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
deconstructed, and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\todo[inline]{Tens de dizer o que significa o refinamento concreto e o
que e a funcao mult e add}
\paragraph{Refinements.} Refinements are logical predicates
which must hold for any element that inhabits a given type. The \synname\
language supports simple refinements on integers with arithmetic
expressions in the predicates. They assert the robustness of our main
framework, i.e., how it is amenable to significant additions without
necessarily interfering with the overall approach.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
synthetised.

\paragraph{Guiding the Search.} Specifications can be additionally
augmented with three keywords: ``using'', ``depth'' and
``assert''. Through them, we can fine-tune the synthesis process and,
respectively, force certain functions to be present in the synthesis
outcome, allow a \emph{deeper} search in the valid programs space, and
require given predicates to hold true in the program after the
synthesis is complete.

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
more intricate example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we synthesize a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: ((a -o b -o a) -> a -o (List b) -o a);
newMArray :: (Int -> (MArray a -o !b) -o b);
write :: (MArray a -o (Int * a) -> MArray a);
freeze :: (MArray a -o !(Array a));
synth array :: (Int -> List (!(Int * a)) -> Array a)
    | using (foldl) | depth 3;
\end{minted}
%
Matching exactly the linear definition for
\emph{array} from~\cite{Bernardy_2018}, \synname\
outputs (in a fraction of a second):
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

@@

\todo[inline]{Indicar o judgment ``declarativo'' e depois os
  judgments em focusing}

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a
system of resource management and focusing.
%
In intuitionistic sequent calculi, each connective has a so-called
\emph{left} and a \emph{right} rule, which effectively define how to
decompose an ambient assumption of a given proposition and how to
prove a certain proposition is true, respectively. 
Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible
and non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Synchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Asynchronous connectives are dual. \todo{Say which
  are which}
%
Given this separation, focusing divides proof search into two
phases: %
the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
to focus on, and then apply non-invertible rules, staying \emph{in focus}
until we reach an asynchronous (i.e. invertible proposition), the
proof is complete or no rules are applicable, in which the proof must
\emph{backtrack} to the state at which the focusing phase began.

\todo[inline]{Isto so faz sentido olhando para os judgments}
For the invertible rules, a
third, ordered, context ($\Omega$) is needed to hold propositions
resulting from decomposing connectives; propositions we'll later try
to invert, moving them to the linear context ($\Delta$) when we fail
to.

@@

\todo[inline]{A frente de cada ``descricao'', escreve o tipo pois
  ajuda o leitor}
The core language is a simply-typed linear lambda calculus with linear
functions, additive and multiplicative pairs, multiplicative unit, additive sums
and the exponential modality (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (heavily inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
first rule \todo{qual ``first''? referencia por nome} is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

A more general formulation of ADTs says an ADT can be recursive (or
"inductively defined"), meaning constructors can take as arguments
values of the type they are defining. This change has
a significant impact in the synthesis process. Take, for instance, the ADT
defined as \mintinline{haskell}{data T = C1 T}, the synthesis goal
$T \lolli C$, and its derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times
the left ADT rule, never closing the proof. Symmetrically, the
derivation for the goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {\dots}
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, symmetrically, don't allow construction of an
ADT when trying to synthetise an argument for its constructor.
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. Together, they hold a list of ADTs that cannot be constructed or
deconstructed at a given point in the proof. For convenience, they are represented by a single $\Rho$ if
unused. All non-ADT rules trivially propagate these. The ADT rules
are then extended as follows, where $\Rho'_C = \Rho_C,T$ if $T$ is
recursive and $\Rho'_C = \Rho_C$ otherwise (similarly for $\Rho'_D$):
%
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
% where
% \begin{mathpar}
%     \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
%     T\ \textrm{\textbf{else}}\ \Rho_C

@@

\mypara{Type-and-Example-Directed Program Synthesis} This
work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
explores a purely functional, recursive program synthesis technique
based on types/proof search (as we intend to do), with examples as an
auxiliar technique to trim down the program synthesis vast search
space. A good use is
made of the general strategy of using a type system to generate
programs (by ``inverting'' the type system), rather than using it to
type-check one.  Moreover, the paper uses a data structure called
``refinement tree'' in conjunction with the extra information provided
by the examples to allow for efficient synthesis of non-trivial
programs. Overall it employs good engineering to achieve type-based
synthesis of functional typed programs, with at least two additions
that we might want to follow too (type refinements and recursive
functions).

>>>>

Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve programmer productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms such as natural
language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
of specification, program synthesis must deal with two main
inherent sources of complexity -- search over the space of
valid programs, and interpreting user intent.

Synthesis is said to be \emph{type-driven} when it uses types as a form of program
specification and produces an expression whose type matches the specification.
Type-driven synthesis frameworks usually leverage rich types as a way to make
specifications more expressive and prune the valid programs search space, while
maintaining a ``familiar'' specification interface (types) for the user.
%
Richer type systems allow for more precise types, which can statically eliminate
various kinds of logical errors by making certain invalid program states
ill-typed (e.g.,~a ``null aware'' type system will ensure at compile-time that
you cannot dereference a null-pointer).
%
For instance, the type $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow
\mathsf{Int}$ specifies a (curried) function that takes two integers and
produces an integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).  However,
the richer type $(x{:}\mathsf{Int}) \rightarrow (y{:}\mathsf{Int}) \rightarrow
\{z{:}\mathsf{Int} \mid z = x+y\}$ very precisely specifies a function that
takes integer arguments $x$ and $y$ and returns an integer $z$ that is the sum
of $x$ and $y$.

The focus of our work is on type-driven synthesis where specifications take the
form of linear types.
Linear types constrain resource
usage in programs by \emph{statically} limiting the number of times certain
resources can be used during their lifetime (linear resources must be used
\emph{exactly once}).
%
They can be applied to resource-aware programming such as concurrent programming
(e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
(e.g.~Rust's ownership types), safely updating-in-place mutable
structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
% relacionar o nosso trabalho com futuras aplicações nestes campos

\paragraph{Contributions and Outline.} Despite their long-known potential
\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
and strong proof-theoretic
foundations
\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
synthesis with linear types combined with other advanced typing features
has generally been overlooked in the literature.  In this work we present a framework for synthesis
with linear types extended with recursive algebraic data types, parametric
polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?
We first introduce linear types as specifications
and outline the synthesis process, leveraging linearity, by example
(\S~\ref{sec:overview}). We then present the formal system driving the
synthesis (\S~\ref{sec:formal_system}) and describe the architecture
of our framework named \synname, examining technical details and key
implementation challenges (\S~\ref{sec:architecture}). Finally, 
\todo{como formular a frase sobre o apêndice?}
we evaluate our work through expressiveness and performance benchmarks
(\S~\ref{sec:evaluation}) and discuss related work
(\S~\ref{sec:related}).

@@

\paragraph{Core linear calculus.} Initially, we synthesize from
specifications that only use literals and propositional linear types, such as linear logic theorems. Namely,
the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
the function \mintinline{haskell}{\a -> \b -> let c*d = b in a c d},
which takes an argument \mintinline{haskell}{a}, a
\emph{curried} linear function of type $A \lolli B \lolli C$,  and an
argument \mintinline{haskell}{b}, a \emph{linear
  pair} of an $A$ and a $B$ and produces a value of type $C$ by
deconstructing the pair and applying \mintinline{haskell}{a} to the
corresponding elements of the pair (i.e.,~\emph{uncurries} the
function \mintinline{haskell}{a}).

@@

\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring
some types to be used linearly (consumed), we can assure, e.g., the
deconstruction of an ADT. As such, the specification that identifies a function
that takes as arguments an unrestricted ($\bang$) linear function\footnote{might be used non-linearly, but its parameter is used
linearly in its body} 
($\lolli$) that converts $a$s in $b$s, a list of $a$s, and produces a
list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthetise
a valid program from it, \mintinline{haskell}{(List a)} has to be deconstructed,
and its constructor arguments used. \synname\ would output the following
function (\emph{map}) without relying on any additional information:
%
% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
% visto que falo dos outros temas? se calhar faço um à parte?
%
\begin{minted}{haskell}
map c d = let !e = c in
  case d of
      Nil -> Nil
    | Cons f -> let g*h = f in Cons (e g, map (!e) h);
\end{minted}

\paragraph{Refinements.} Refinements are logical predicates
which must hold for any element that inhabits a given type. The \synname\
language supports simple refinements on integers with arithmetic
expressions in the predicates. They assert the robustness of our main
framework, i.e., how it is amenable to significant additions without
necessarily interfering with the overall approach.
%
As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$ (a function that takes three
integers and produces an integer that satisfies the given predicate), the
program\ \mintinline{haskell}{(\a -> \b -> \c -> ((0 - 1) * a) + (c * b))} is
synthetised.

\paragraph{Guiding the Search.} Specifications can be additionally
augmented with four keywords: ``using'', ``depth'', ``assert'' and ``choose''.
Through them, we can fine-tune the synthesis process and,
respectively, force certain functions to be present in the synthesis
outcome, allow a \emph{deeper} search in the valid programs space,
require given predicates to hold true in the program after the
synthesis is complete, and select between alternative results.

Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
more intricate example of synthesis: given the linear type signatures for array
primitives (\emph{newMArray} passes a new mutable array to a function that uses
it linearly, \emph{write} takes a mutable array and writes a pair to it,
\emph{freeze} consumes a mutable array and produces an imutable array,
\emph{foldl} has the default definition) we synthesize a function
``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
follows:
%
\begin{minted}{haskell}
foldl :: (a -o b -o a) -> a -o (List b) -o a;
newMArray :: Int -> (MArray a -o !b) -o b;
write :: MArray a -o (Int * a) -> MArray a;
freeze :: MArray a -o !(Array a);
synth array :: Int -> List (!(Int * a)) -> Array a
    | using (foldl) | depth 3;
\end{minted}
%
Matching exactly the linear definition for
\emph{array} from~\cite{Bernardy_2018}, \synname\
outputs (in a fraction of a second):
%
\begin{minted}{haskell}
array b c = let !d = b in let !e = c in
    newMArray (!d) (\j -> freeze (foldl (!write) j e));
\end{minted}

@@

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a system of
resource management~\cite{resman} and focusing.
%
% TODO: Onde falamos de resource management? ( 
%
In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Asynchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Synchronous connectives are dual. \todo{Say which
  are which}

Given this separation, focusing divides proof search into two
phases: %
the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
to focus on, and then apply non-invertible rules, staying \emph{in focus}
until we reach an asynchronous (i.e. invertible proposition), the
proof is complete, or no rules are applicable, in which case the proof must
\emph{backtrack} to the state at which the focusing phase began.
As such, with focusing, a linear logic judgment in sequent calculus ($\Gamma; \Delta, B
\vdash A$) is broken down into four different judgments, grouped into the two
phases ($\Uparrow, \Downarrow$). For the invertible phase, a third % TODO: ordered??,
context ($\Omega$) is needed to hold propositions that result from
decomposing connectives (propositions we'll later try to invert, moving them to
the linear context ($\Delta$) when we fail to). The judgments of right
invertible rules are written $\Gamma; \Delta; \Omega,B \vdash A \Uparrow$, and
those of left invertible rules are $\Gamma; \Delta; \Omega,B \Uparrow\ \vdash A$.
In the focusing phase, the focus proposition will be either the goal or come
from $\Gamma$ or $\Delta$, and appliable rules will be classified as right
non-invertible, or left non-invertible, respectively. Right non-invertible
rules have the form $\Gamma;\Delta,B \vdash A \Downarrow$, and left
non-invertible rules are written $\Gamma;\Delta; B \Downarrow\ \vdash A$.

@@

The core language is a simply-typed linear lambda calculus with linear
functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
multiplicative unit (\textbf{1}), additive sums ($\oplus$)
and the exponential modality ($\bang$) (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (%heavily
inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
rule ($\lolli R$) is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

A more general formulation of ADTs says an ADT can be recursive (or
"inductively defined"), meaning constructors can take as arguments
values of the type they are defining. This change has
a significant impact in the synthesis process. Take, for instance, the ADT
defined as \mintinline{haskell}{data T = C1 T}, the synthesis goal
$T \lolli C$, and its derivation:
%
\begin{mathpar}
    \infer*[right=($\lolli R$)]
    {
        \infer*[right=($\Uparrow R$)]
        {
            \infer*[right=(adtL)]
            {
                \infer*[right=(adtL)]
                {\dots}
                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
                \textrm{of}\ C_1\ z \rightarrow \dots : C}
            }
            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
            \textrm{of}\ C_1\ y \rightarrow \dots : C}
        }
        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
        \Uparrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
    T \lolli C \Uparrow}
\end{mathpar}
%
Using our current system, we are to apply an infinite number of times
the left ADT rule, never closing the proof. Symmetrically, the
derivation for the goal $T$ is also infinite.
%
\begin{mathpar}
    \infer*[left=(adtR)]
    {
        \infer*[left=(adtR)]
        {
            \infer*[left=(adtR)]
            {\dots}
            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
        }
        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
    }
    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
\end{mathpar}
%
To account for this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and, symmetrically, don't allow construction of an
ADT when trying to synthetise an argument for its constructor.
%
%
For this, we need two more contexts, $\Rho_C$
for constraints on construction and $\Rho_D$ for constraints on
deconstruction. Together, they hold a list of ADTs that cannot be constructed or
deconstructed at a given point in the proof. For convenience, they are represented by a single $\Rho$ if
unused. All non-ADT rules trivially propagate these. The ADT rules
are then extended as follows, where $\Rho'_C = \Rho_C,T$ if $T$ is
recursive and $\Rho'_C = \Rho_C$ otherwise ($\Rho'_D$ is dual):
%
%TODO: ... we can actually instance ADTs that take no arguments even if they are restricted
\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    T \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        T \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
% where
% \begin{mathpar}
%     \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
%     T\ \textrm{\textbf{else}}\ \Rho_C

@@

% \mypara{Type-and-Example-Directed Program Synthesis} This
% work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
% explores a purely functional, recursive program synthesis technique
% based on types/proof search (as we intend to do), with examples as an
% auxiliar technique to trim down the program synthesis vast search
% space. A good use is
% made of the general strategy of using a type system to generate
% programs (by ``inverting'' the type system), rather than using it to
% type-check one.  Moreover, the paper uses a data structure called
% ``refinement tree'' in conjunction with the extra information provided
% by the examples to allow for efficient synthesis of non-trivial
% programs. Overall it employs good engineering to achieve type-based
% synthesis of functional typed programs, with at least two additions
% that we might want to follow too (type refinements and recursive
% functions).
